---
title: "Mark-recapture model verification"
author: "Maxwell B. Joseph and Roland A. Knapp"
date: "4/10/2018"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Simulating data with known parameters with the goal of evaluating parameter recovery. 

```{r load-deps, message=FALSE}
library(tidyverse)
library(reshape2)
library(assertthat)
library(rstan)
set.seed(1234)
```

## Generating sampling design quantities

First, we will determing the number of primary periods $T$, and the snow water content for each primary period, which will be used as a covariate. 

```{r make-primary-periods}
# fixed data
T <- 40 # number of primary periods
Tm1 <- T - 1
K <- 15 # number of years
year <- round(seq(1, K, length.out = T))
is_overwinter <- c(diff(year), 0)

# covariate value (snow water content) for primary periods
x <- rnorm(K) # yearly values
x_t <- x[year] # values by primary period
x_tm1 <- c(rnorm(1), x_t[1:T-1])

```

Next, we will determine the number of secondary periods (surveys) that take place for each primary period. 
By design, there are no surveys for the first primary period. 

```{r make-secondary-periods}
# rpois() allows for primary periods with 0 secondary periods
# essentially a missed, unsurveyed primary period
J <- c(0, rpois(T - 1, lambda = 4))
Jtot <- sum(J)

# create a vector of indices for each secondary period
jvec <- c()
for (i in seq_along(J)) {
  if (J[i] > 0) {
    jvec <- c(jvec, 1:J[i])
  }
}

stopifnot(length(jvec) == Jtot)
```

Now, we determine the total number of adults that are introduced, the total number of individuals ever to have existed $M$, and determine when each introduced animal enters the population. 

```{r specify-introduced-adults}
n_intro <- 20
M <- n_intro + 380 # augmented num indiv
introduced <- c(rep(TRUE, n_intro), rep(FALSE, M - n_intro))


# determine which primary period each individual was introduced
# must be 2 or more, because the 1st primary period is make-believe
t_intro <- sample(2:Tm1, size = n_intro, replace = TRUE) %>%
  sort %>%
  c(rep(0, M - n_intro))
```

## Bd load data

Next, we will determine the average log Bd load among infected adults for each primary period, which is modeled as a normally distributed random variable whose mean is a linear function of snow water content. 

```{r specify-mean-load}
# determine average intensity for infected individuals at each
# primary period
alpha_z <- 0
beta_z <- 1
sigma_z <- .5
mu_z <- rnorm(n = K, mean = alpha_z + beta_z * x, sd = sigma_z)
```

Using these average values, we can determine the potential log loads for each adult at each primary period, conditional on infection. 

```{r specify-potential-load}
# individual variation around mean load
sigma_bd_indiv <- .2

potential_log_load <- matrix(nrow = M, ncol = T)

# load conditional on being infected
for (i in 1:T) {
  potential_log_load[, i] <- rnorm(n = M, mean = mu_z[year[i]], sd = sigma_bd_indiv)
}
```

## Detection quantities

Next, we need to determine survey-level covariates (e.g., air temperature) and use these to generate a detection design matrix. 

```{r}
n_p <- 2 # number of columns in detection design matrix
assert_that(n_p >= 2)

X_p <- cbind(rep(1, Jtot), 
             matrix(rnorm(Jtot * (n_p - 1)), ncol = n_p - 1))
```

## Parameter generation

Next, specify the remaining true parameter values:

```{r}
# recruitment parameters
alpha_lambda <- -4
beta0_lambda <- 0
beta_lambda_ow <- rnorm(1)
beta_lambda_wc <- rnorm(1)
sigma_lambda <- 1.2
eps_lambda <- rnorm(K, 0, sigma_lambda)

# survival parameters: uninfected adults
alpha_phi_uninfected <- 3
beta_phi_ow <- rnorm(1)
beta_phi_wc <- rnorm(1)
sigma_phi_uninf <- abs(rnorm(1))
eps_phi_uninf <- rnorm(K, 0, sigma_phi_uninf)

# survival parameters: infected adults
alpha_phi_infected <- 2.5
beta_phi_inf_wc <- rnorm(1)
beta_phi_infected <- rnorm(1)
beta_phi_inf_ow <- rnorm(1)
sigma_phi_inf <- abs(rnorm(1))
eps_phi_inf <- rnorm(K, 0, sigma_phi_inf)

# loss of infection
alpha_eta1 <- rnorm(1)
beta_eta1 <- rnorm(1)
sigma_eta1 <- abs(rnorm(1))
eps_eta1 <- rnorm(K, 0, sigma_eta1)

# gain of infection
alpha_eta2 <- rnorm(1)
beta_eta2 <- abs(rnorm(1))
sigma_eta2 <- abs(rnorm(1))
eps_eta2 <- rnorm(K, 0, sigma_eta2)

# conditional probability of being infected, given recruitment
alpha_g <- rnorm(1)
beta_g <- rnorm(1)
sigma_g <- abs(rnorm(1))
eps_g <- rnorm(K, 0, sigma_g)

# detection parameters
beta_p <- c(0, rnorm(n_p - 1))
beta_p_infected <- rnorm(1)
beta_p_bd <- rnorm(1)

# probability that a swab succeeds
delta <- .9
```


## Specifying transition probabilities

We'll begin specifying transition matrices with survival probabilities:

```{r}
phi_infected <- array(dim = c(M, T))
phi_uninfected <- array(dim = c(M, T))
for (t in 1:T) {
  phi_uninfected[, t] <- plogis(alpha_phi_uninfected +
                                beta_phi_ow * is_overwinter[t] +
                                beta_phi_wc * x_t[t] +
                                eps_phi_uninf[year[t]])

  phi_infected[, t] <- plogis(alpha_phi_infected + 
                              beta_phi_infected * potential_log_load[i, t] + 
                              beta_phi_inf_ow * is_overwinter[t] + 
                              beta_phi_inf_wc * x_t[t] + 
                              eps_phi_inf[year[t]])
}
```

Then, probabilities of entering the population:

```{r}
lambda <- alpha_lambda + 
          beta_lambda_ow * is_overwinter[1:Tm1] + 
          beta_lambda_wc * x_t[1:Tm1] + 
          eps_lambda[year[1:Tm1]]
lambda[1] <- lambda[1] + beta0_lambda
lambda <- plogis(lambda)
```

Probability of recruiting as infected, given entry:

```{r}
g <- plogis(alpha_g + 
            beta_g * mu_z[year[1:Tm1]] + 
            eps_g[year[1:Tm1]])

gamma <- matrix(nrow = 2, ncol = Tm1)
gamma[1, ] <- lambda * (1 - g)   # recruit, not infected
gamma[2, ] <- lambda * g         # recruit, infected
```

Loss and gain of infection:

```{r}
# loss of infection
eta1 <- plogis(alpha_eta1 + 
               beta_eta1 * mu_z[year[1:Tm1]] + 
               #beta_eta1_ow * is_overwinter[1:Tm1] + 
               eps_eta1[year[1:Tm1]])

# gain of infection
eta2 <- plogis(alpha_eta2 +
               beta_eta2 * mu_z[year[1:Tm1]] + 
               #beta_eta2_ow * is_overwinter[1:Tm1] + 
               eps_eta2[year[1:Tm1]])
```

Now, we can use these quantities to generate transition matrices. 
We will store these in an array `ps` with dimensions corresponding to starting states, individuals, primary periods, and destination states. 

```{r}
ps <- array(dim = c(4, M, Tm1, 4))
for (i in 1:M) {
  # fill in shared values
  for (t in 1:Tm1) {
    ps[1, i, t, 4] <- 0                                      # can't die before being alive
    ps[2, i, t, 1] <- 0                                      # can't unenter population
    ps[3, i, t, 1] <- 0
    ps[2, i, t, 2] <- phi_uninfected[i, t] * (1 - eta2[t])   # survive/didn't get infected
    ps[2, i, t, 3] <- phi_uninfected[i, t] * eta2[t]         # survived/became infected
    ps[2, i, t, 4] <- 1 - phi_uninfected[i, t]               # death
    ps[3, i, t, 2] <- phi_infected[i, t] * eta1[t]           # survived, lost infection
    ps[3, i, t, 3] <- phi_infected[i, t] * (1 - eta1[t])     # survived, didn't lose infection
    ps[3, i, t, 4] <- 1 - phi_infected[i, t]                 # death
    ps[4, i, t, 1] <- 0                                      # can't unenter after death
    ps[4, i, t, 2] <- 0                                      # can't un-die
    ps[4, i, t, 3] <- 0
    ps[4, i, t, 4] <- 1                                      # once dead, always dead
  }
  
  if (introduced[i]) {
    # zero probability of recruiting through t_intro - 2
    for (t in 1:(t_intro[i] - 2)) {
      ps[1, i, t, 1] <- 1
      ps[1, i, t, 2] <- 0
      ps[1, i, t, 3] <- 0
    }
    # timestep before introduction has Pr(recruiting) = 1
    # NOTE: this assumes introduced animals are infected upon introduction
    ps[1, i, t_intro[i] - 1, 1] <- 0
    ps[1, i, t_intro[i] - 1, 2] <- 0
    ps[1, i, t_intro[i] - 1, 3] <- 1
    # to avoid NA values, fill in remaining recruitment probs (though they
    # are irrelevant for the likelihood)
    for (t in t_intro[i]:Tm1) {
      ps[1, i, t, 1] <- 1 - gamma[1, t] - gamma[2, t]
      ps[1, i, t, 2] <- gamma[1, t]
      ps[1, i, t, 3] <- gamma[2, t]
    }
  } else {
    # they have not been introduced, so they must recruit naturally
    for (t in 1:Tm1) {
      ps[1, i, t, 1] <- 1 - gamma[1, t] - gamma[2, t]
      ps[1, i, t, 2] <- gamma[1, t]
      ps[1, i, t, 3] <- gamma[2, t]
    }
  }
}

# verify no NA values in transition probabilities
assert_that(!any(is.na(ps)))

# verify that all transition matrices are simplexes
for (i in 1:M) {
  for (t in 1:Tm1) {
    # rows must sum to one
    assert_that(all(abs(rowSums(ps[, i, t, ]) - 1) < 1e16))
  }
}
```

## Simulating true states

Now that we have the transition probabilities, we can simulate the actual states. 

```{r}
# Simulating true states --------------------------------------------------
z <- array(dim = c(M, T))
load <- matrix(nrow = M, ncol = T)

z[, 1] <- 1               # all animals are assumed to be in state 1 initially
for (i in 1:M) {
  for (t in 2:T) {
    z[i, t] <- which.max(rmultinom(1,
                                   size = 1,
                                   prob = ps[z[i, t - 1], i, t - 1, ]))
    if (z[i, t] == 3) {
      load[i, t] <- exp(potential_log_load[i, t])
    }
    if (z[i, t] == 2) {
      load[i, t] <- 0
    }

  }
}

rownames(z) <- paste0('individual_', 1:M)
colnames(z) <- paste0('primary_period_', 1:T)

rownames(load) <- rownames(z)
colnames(load) <- colnames(z)

# assert no missing values
assert_that(!any(is.na(z)))

# confirm that some of these individuals never existed (there are pseudo-individuals)
pseudoindividuals <- rowSums(z) == T
assert_that(any(pseudoindividuals))
```

Now, let's visualize the states. 

```{r}
z %>%
  as.data.frame %>%
  tbl_df %>%
  rownames_to_column %>%
  gather(primary_period, state, -rowname) %>%
  mutate(primary_period = parse_number(primary_period), 
         individual = parse_number(rowname), 
         introduced = introduced[individual]) %>%
  group_by(individual) %>%
  mutate(first_recruited = primary_period[state > 1][1]) %>%
  ungroup %>%
  mutate(state = case_when(
    .$state == 1 ~ 'Not recruited', 
    .$state == 2 ~ 'Uninfected', 
    .$state == 3 ~ 'Infected', 
    .$state == 4 ~ 'Dead'
  )) %>%
  ggplot(aes(x = primary_period, 
             y = reorder(factor(individual), first_recruited), 
             fill = state)) + 
  geom_tile() + 
  scale_fill_manual(values = c('black', 'red2', 'lightgrey', 'dodgerblue')) + 
  theme(axis.text.y = element_blank(), 
        axis.title.y = element_blank())
```

## Simulating observations

Now that we have the states, we can simulate observations. 

```{r}
po <- array(0, dim = c(4, M, Jtot, 4))
prim_idx <- rep(1:T, J)

logit_p <- X_p %*% beta_p

for (i in 1:M) {
  for (j in 1:Jtot) {
    p_uninf <- plogis(logit_p[j])
    p_inf <- plogis(logit_p[j] + 
                      beta_p_infected + 
                      beta_p_bd * potential_log_load[i, prim_idx[j]])
  
    po[1, i, j, 1] <- 1
    po[1, i, j, 2] <- 0
    po[1, i, j, 3] <- 0
    po[1, i, j, 4] <- 0

    po[2, i, j, 1] <- 1 - p_uninf
    po[2, i, j, 2] <- p_uninf * delta
    po[2, i, j, 3] <- 0
    po[2, i, j, 4] <- p_uninf * (1 - delta)

    po[3, i, j, 1] <- 1 - p_inf
    po[3, i, j, 2] <- 0
    po[3, i, j, 3] <- p_inf * delta
    po[3, i, j, 4] <- p_inf * (1 - delta)

    po[4, i, j, 1] <- 1
    po[4, i, j, 2] <- 0
    po[4, i, j, 3] <- 0
    po[4, i, j, 4] <- 0
    assert_that(all(abs(rowSums(po[, i, j, ]) - 1) < 1e12))
  }
}
```

Given the true states and the observation matrices, simulate observations:

```{r}
# simulate detection data
y_prim <- rep(1:T, J)
y <- array(dim = c(M, Jtot))
for (i in 1:M) {
  for (j in 1:Jtot) {
    y[i, j] <- which.max(rmultinom(1, 1, po[z[i, y_prim[j]], i, j, ]))
  }
}

# create a survey number data frame to ensure each primary period is represented
# in the observation array Y, even if there were no surveys, simplifying indexing
survey_number_df <- tibble(primary_period = 1:T) %>%
  mutate(secondary_idx = 1, individual = 1)

y_df <- melt(y,
             varnames = c('individual', 'secondary_period'),
             value.name = 'y') %>%
  tbl_df %>%
  mutate(primary_period = y_prim[secondary_period],
         secondary_idx = jvec[secondary_period]) %>%
  dplyr::select(individual, primary_period, secondary_idx, y) %>%
  full_join(survey_number_df)

Y <- acast(y_df, individual ~ primary_period ~ secondary_idx, fill = 0,
           value.var = "y")

# check dimension match for Y
assert_that(dim(Y)[3] == max(J))
```

In addition to the observations, we need to have indices for the order of surveys:

```{r}
p_df <- tibble(primary_period = rep(1:T, J),
               secondary_idx = jvec) %>%
  mutate(j_idx = 1:n())


j_idx <- survey_number_df %>%
  dplyr::select(primary_period, secondary_idx) %>%
  full_join(p_df) %>%
  acast(primary_period ~ secondary_idx,
        fill = 0,
        value.var = 'j_idx')

# verify that the primary periods with no surveys have all zeros in j_idx
assert_that(identical(names(which(rowSums(j_idx) == 0)),
            which(J == 0) %>% as.character))
```

Now that we have the actual observations, we can determine which swabs would have been collected to get a swab dataset. 

```{r}
# load_df is an oracle data frame that has all of the infected animal
# infection intensities
load_df <- melt(load,
                varnames = c('individual', 't'),
                value.name = 'z') %>%
  as_tibble %>%
  mutate(t = parse_number(t),
         individual = parse_number(individual))

# subset y_df to figure out which swabs would have been collected and run
swab_df <- y_df %>%
  filter(y == 2 | y == 3) %>%
  group_by(individual, primary_period) %>%
  summarize(obs_state = unique(y)) %>%
  ungroup() %>%
  rename(t = primary_period) %>%
  left_join(load_df)

pos_swab_df <- filter(swab_df, z > 0)

pos_swab_df %>%
  ggplot(aes(t, z)) + 
  geom_jitter(height = 0, width = .2) + 
  xlab('Primary period') + 
  ylab('Observed load')
```

Having identified all of the positive swabs, we can proceed to create the array of potential log loads:

```{r}
load_observed <- matrix(0, nrow = M, ncol = T)
observed_log_load <- matrix(-9999, nrow = M, ncol = T)
for (i in 1:M) {
  for (prim_period in 1:T) {
    swab_subset <- filter(swab_df, individual == i & t == prim_period)
    if (nrow(swab_subset) > 0) {
      if (swab_subset$z != 0) {
        # we are only tracking positive (nonzero) swab values
        load_observed[i, prim_period] <- 1
        observed_log_load[i, prim_period] <- log(swab_subset$z)
      }
    }
  }
}
```

Finally, we can bundle the data to use in our Stan model: 

```{r}
stan_d <- list(M = M,
               T = T,
               maxJ = max(J),
               J = J,
               Jtot = Jtot,
               Y = Y,
               introduced = introduced,
               t_intro = t_intro,
               X_p = X_p,
               n_p = ncol(X_p),
               j_idx = j_idx,
               any_surveys = ifelse(J > 0, 1, 0),
               n_swab = nrow(pos_swab_df),
               t_swab = pos_swab_df$t,
               log_load = log(pos_swab_df$z),
               load_observed = load_observed,
               observed_log_load = observed_log_load,
               is_overwinter = is_overwinter,
               prim_idx = rep(1:T, J),
               x = x,
               x_t = x_t,
               x_tm1 = x_tm1,
               K = K,
               year = year, 
               different_detection_prob = 1)
```

Then, let's run ADVI to estimate parameters. 

```{r}
m_init <- stan_model('../stan/uncertain-state.stan')
m_fit <- vb(m_init,
            data = stan_d,
            eta = .1, tol_rel_obj = 0.005,
            init = 0,
            adapt_engaged = FALSE)
```

```{r}
post <- rstan::extract(m_fit)

# check to see whether the 
hist(post$Nsuper)
true_Nsuper <- M - sum(pseudoindividuals)
abline(v = true_Nsuper, col = 2, lty = 2)
```

Now, let's evaluate the correspondence between the true and estimated abundance. 

```{r}
par_names <- c('alpha_lambda',
               'beta0_lambda', 'beta_lambda_ow', 
               'beta_lambda_wc',  
               'sigma_lambda',
               'alpha_phi_uninfected', 'beta_phi_ow', 
               'beta_phi_wc', 
               'sigma_phi_uninf', 
               'alpha_phi_infected', 'beta_phi_inf_wc', 
               'beta_phi_infected', 
               'beta_phi_inf_ow', 'sigma_phi_inf',
               'alpha_eta1', 'beta_eta1',  
               'sigma_eta1',
               'alpha_eta2', 'beta_eta2', 
               'sigma_eta2', 
               'alpha_g', 'beta_g', 
               'sigma_g',
               'beta_p', 'beta_p_infected', 'beta_p_bd', 
               'delta')

qvals <- c(0.025, .5, 0.975)
estimates <- lapply(post[par_names], quantile, qvals) %>%
  bind_rows(.id = 'par')
estimates$par <- qvals
estimates <- estimates %>%
  gather(parameter, value, -par) %>%
  as_tibble
estimates$true <- sapply(estimates$parameter, FUN = function(x) eval(parse(text = x))[1]) %>% 
  unlist
estimates <- estimates %>%
  spread(par, value) %>%
  mutate(family = case_when(
    grepl('eta1', .$parameter) ~ 'Loss of infection', 
    grepl('eta2', .$parameter) ~ 'Gain of infection', 
    grepl('lambda', .$parameter) ~ 'Recruitment',
    grepl('phi', .$parameter) ~ 'Survival',
    grepl('_g', .$parameter) ~ 'Conditional infection', 
    TRUE ~ 'Other'
  ))
  

estimates %>%
  ggplot(aes(x = true, y = `0.5`)) + 
  geom_point() + 
  geom_linerange(aes(ymin = `0.025`, ymax = `0.975`)) + 
  geom_abline(slope = 1, intercept = 0, linetype = 'dashed') + 
  xlab('True value') + 
  ylab('Estimate with approximate 90% credible interval') + 
  facet_wrap(~family)
```

Last, let's make a table to evaluate parameter recovery for this realization:

```{r}
estimates %>% 
  arrange(family) %>%
  knitr::kable()
```
